---
title: "Calculating summary values over bigWig files"
author: "Carmen Navarro"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{Value calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{wigglescout}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Summary

`wigglescout` allows you to summarize bigWig files in two main ways:

- **Genome-wide**. Genome is partitioned on equally-sized bins and
their aggregated value is calculated. Useful to get a general idea of the
signal distribution without looking at specific places.
- **Across a set of *loci***. This can be either summarized categories, or
individual values, as in genome-wide analyses.

`wigglescout` functionality is built in two layers. Names of functions that
calculate values over bigWig files start with `bw_`. These return `GRanges`
objects when possible, `data.frame` objects otherwise (i.e. when values are
summarized over some category, genomic location is lost in this process).

# Available functions

bigWig data can be summarized genome-wide using `bw_bins` function or across
a set of *loci* using either: `bw_bed`, `bw_heatmap` or `bw_profile`.


# About bundled data

This package comes with a set of small files to show functionality. These
have been built from a published data from Simon Elsässer's lab and correspond
to a H3.3 and H3K9me3 ChIP + input (`GSE149080`), subset across a
500kbp genomic region: `chr15-102600000-103100000`, which overlaps with the
*HOXC* gene cluster. A ChromHMM annotation has also been subset to overlap 
with such region.

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(scales)
library(rtracklayer)
library(GenomicRanges)
library(wigglescout)

h33_chip <- system.file("extdata", "sample_H33_ChIP.bw", package = "wigglescout")
h3k9me3_chip <- system.file("extdata", "sample_H3K9me3_ChIP.bw", package = "wigglescout")
input_chip <- system.file("extdata", "sample_Input.bw", package = "wigglescout")
genes <- system.file("extdata", "sample_genes_mm9.bed", package = "wigglescout")
chromhmm <- system.file("extdata", "sample_chromhmm.bed", package = "wigglescout")

locus <- GRanges(seqnames = "chr15", IRanges(102600000, 103100000))
```

All these values are paths to bigWig and BED files.

# Genome-wide analysis

`bw_bins` function returns a `GRanges` object where each entry is a locus of 
length `bin_size` and its score is the mean coverage.

**Note:** Small bin sizes (< 5000bp) take longer to run.

```{r, warning = FALSE, message = FALSE}
# Several bwfiles can be provided at once
bw_bins(c(h33_chip, input_chip),
        bin_size = 50000,
        genome = "mm9",
        selection = locus)
```

It is also possible to provide a list of bigWig files to be used as background
to normalize the bin values to. For instance, in the previous case, one could
want to use the input values to normalize the H3.3 ChIP data:

```{r, warning = FALSE, message = FALSE}
bw_bins(h33_chip, bg_bwfiles = input_chip,
        bin_size = 50000,
        genome = "mm9",
        selection = locus)
```
If `log2` is passed as `norm_func` parameter, we get the log2 ratio sample / input:

```{r, warning = FALSE, message = FALSE}
bw_bins(h33_chip, bg_bwfiles = input_chip,
        bin_size = 50000,
        genome = "mm9",
        selection = locus,
        norm_func = log2) # Note this is not a string, but the actual function
```

# Locus-based analysis

These functions are useful when you have some genomic annotations you want to
look at, rather than general genome-wide trends.

## Individual loci

To calculate values on a *locus* specific manner you can use `bw_bed` function.
For instance, we can look at the genes at this region:

```{r, warning = FALSE, message = FALSE}
bw_bed(c(h33_chip, input_chip), bedfile = genes)
```

Many of the options available for `bw_bins` are shared with `bw_bed`, so you can
also use background bigWig files:

```{r, warning = FALSE, message = FALSE}
bw_bed(h33_chip, bg_bwfiles = input_chip, bedfile = genes)
```

A name column is provided if the given `BED` file includes *loci* IDs.

## Aggregated loci

It is also possible to aggregate values by some kind of category. This is 
useful for `BED` files where ID is not an individual entity but rather a 
category, for instance, ChromHMM annotations:

```{r, warning = FALSE, message = FALSE}
chrom_ranges <- import(chromhmm)

chrom_ranges
```

It is possible to get a single value per category using `aggregate_by` parameter.

```{r, warning = FALSE, message = FALSE}
# If aggregate_by is not provided, you get a non-aggregated GRanges
bw_bed(h33_chip,
       bg_bwfiles = input_chip,
       bedfile = chromhmm,
       aggregate_by = "mean") 
```

**Note:** In this case you will not get a `GRanges` object but a `data.frame`.

For these calculations you **always** get some kind of **average** per locus. 
However, different ways of treating the average values are available by changing
the `aggregate_by` parameter:

- `true_mean`: This calculates a mean adding up all the values included per *locus*. 
It is most meaningful when loci represent categories rather than biological
entities.
- `mean`: This calculates a mean-of-means instead, where each *locus* mean value
is calculated and the mean of this distribution of means is returned as a result.
- `median`: This calculates a median-of-means. It is useful to perform an analysis
that is more robust to outliers, but it does not work well with sparse data.

## Profile averages

You may be interested in the **shape** the signal takes across a set of *loci*.
This is calculated by `bw_profile`. This function takes each *locus* in the 
provided bed file and slices it into a fixed number of bins (this is determined
by `bin_size` parameter). Each bin is aggregated together across loci.


```{r, warning = FALSE, message = FALSE}
profile <- bw_profile(h33_chip, bedfile = genes, bin_size = 100)

head(profile)
```

**Note:**. `bw_profile` returns data as long format, since it returns more than
one value per data point. You get: mean, standard error, median value per point.
Each point is represented by index. This index determines its bin.

You can also provide `upstream` and `downstream` base pairs values to include
in this calculation.

```{r, warning = FALSE, message = FALSE}
profile <- bw_profile(h33_chip, bedfile = genes, bin_size = 100,
                      upstream = 500, downstream = 500)

head(profile)
```

Additionally, you can use `mode` parameter to define how these locus are aligned
before binning. Available modes are: `start`, `end`, `center` (align the loci around
these points) or `stretch` which anchors features `start` to `end` and stretches
them when length is different.

## Per-locus profiles

This shape created by `bw_profile` function is still an average. So you may be
interested on a per-locus profile value. This is what you get with `bw_heatmap`.
`bw_heatmap` returns a `matrix` where each row is a locus and each column is
a bin. The rest of parameters are analogous to `bw_profile`:

```{r, warning = FALSE, message = FALSE}
mat <- bw_heatmap(h33_chip, bedfile = genes, bin_size = 1000,
                  upstream = 5000, downstream = 5000)

# Note that bw_heatmap returns a list of matrices, so it supports arrays of 
# bwfiles as input as well.
head(mat[[1]])
```

# Using multiple processors

As of version 0.2.0, `wigglescout` core functions support `future` specifications
using `furrr` library. This means you can run the code in multiple R sessions:

```{r, warning = FALSE, message = FALSE}
# You just have to set the plan
library(future)
plan(multisession, workers = 2)

# Then run the functions just the same
bw_bins(c(h33_chip, input_chip),
        bin_size = 50000,
        genome = "mm9",
        selection = locus,
        norm_func = log2) # Note this is not a string, but the actual function
```

It is advised against implementing future planning within library function 
according to [`future` documentation](https://www.rdocumentation.org/packages/future/versions/1.21.0/topics/plan):

>Please refrain from modifying the future strategy inside your packages / functions, i.e. do not call plan() in your code. Instead, leave the control on what backend to use to the end user. This idea is part of the core philosophy of the future framework - as a developer you can never know what future backends the user have access to. Moreover, by not making any assumptions about what backends are available, your code will also work automatically with any new backends developed after you wrote your code.

## How to find the best configuration

Parallelization in `wigglescout` setting is based on the fact that operations
on multiple files are done separately and independently of each other. However,
doing this requires some passing of data back and forth, which means that if 
your bigWig files are large (or the results you get from them, for instance,
big `GRanges` objects), or you don't have many, you may suffer the overhead
without getting any benefit from it.

An easy rule of thumb is: swap from `sequential` (which is the default) to
`multisession` if you are running functions on multiple files at once. It seems
like there is benefit to do this pretty much for any number of files larger 
than one.

However, adding more sessions does not seem to help a lot. Best guess is going
for 2 or 4 workers. With 2 workers you already see a considerable reduction
in time.

My expectation is that `multicore` may be faster. However I have not benchmarked
it, as it does not run within Rstudio.



