---
title: "Wigglescout documentation"
author: "Carmen Navarro"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{Value calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{wigglescout}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Summary and scope

`wigglescout` is an R library that allows you to calculate summary values 
across bigWig files and BED files and visualize them in a genomics-relevant
manner. It is based on broadly used libraries such as `rtracklayer` and
`GenomicRanges`, among others for calculation, and mostly `ggplot2` for
visualization. You can look at the `DESCRIPTION` file to get more information
about all the libraries that make this one possible. 

There are also many other tools whose functionality overlaps a little or
much with `wigglescout`, but there was no single tool that 
included all that I needed. The aim of this library is therefore not to 
replace any of those tools, or to provide a silver-bullet solution to genomics
data analysis, but to provide a comprehensive, yet simple enough set of tools
focused on bigWig files that can be used entirely from the R environment
without switching back and forth across tools.

Other tools and libraries for akin purposes that you may be looking for include:
deepTools, SeqPlots, bwtools, wiggletools, and the list is endless!

`wigglescout` allows you to summarize and visualize the contents of 
bigWig files in two main ways:

- **Genome-wide**. Genome is partitioned on equally-sized bins and
their aggregated value is calculated. Useful to get a general idea of the
signal distribution without looking at specific places.
- **Across sets of *loci***. This can be either summarized categories, or
individual values, as in genome-wide analyses.

`wigglescout` functionality is built in two layers. Names of functions that
calculate values over bigWig files start with `bw_`. These return `GRanges`
objects when possible, `data.frame` objects otherwise (i.e. when values are
summarized over some category, genomic location is lost in this process).

On the other hand, functions that plot such values and that usually make
internal use of `bw_` functions, start with `plot_`.

# About bundled data

This package comes with a set of small files to show functionality. These
have been built from a published data from Simon Elsässer's lab and correspond
to a H3.3 and H3K9me3 ChIP + input (`GSE149080`), subset across a
500kbp genomic region: `chr15-102600000-103100000`, which overlaps with the
*HOXC* gene cluster. A ChromHMM annotation has also been subset to overlap 
with such region.

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(scales)
library(rtracklayer)
library(GenomicRanges)
library(wigglescout)

h33_chip <- system.file("extdata", "sample_H33_ChIP.bw", package = "wigglescout")
h3k9me3_chip <- system.file("extdata", "sample_H3K9me3_ChIP.bw", package = "wigglescout")
input_chip <- system.file("extdata", "sample_Input.bw", package = "wigglescout")
genes <- system.file("extdata", "sample_genes_mm9.bed", package = "wigglescout")
chromhmm <- system.file("extdata", "sample_chromhmm.bed", package = "wigglescout")

locus <- GRanges(seqnames = "chr15", IRanges(102600000, 103100000))
```

All these values are paths to bigWig and BED files.

# Quick start

Here is a small example of what you can do with `wigglescout`. Imagine you want
to take a look at how values are distributed overall in your bigwig files,
genome-wide:

```{r, warning = FALSE, message = FALSE, fig.height=7, fig.width=8}
plot_bw_bins_violin(
  c(h33_chip, h3k9me3_chip),
  bin_size = 50000,
  labels = c("H3.3", "H3K9me3"),
  remove_top = 0.001
)
```

For transparency and reproducibility, plots show relevant underlying values by
default, such as parameters used and calculated values: how many `NA` values were
found, if any, how many points were excluded from the plot due to quantile
cutoff, and so on. Note also that in this case the quantile cutoff seems low,
but it is because top elements are removed according to mean in all samples.

In this example, since the bundled data only has signal in a very small subset
of the bins, you only plot a few of those 54531 points, the rest are `NA` 
values.

You could also be interested in checking out how are certain *loci*  behaving
in comparison with the global distribution. You can do so providing a `highlight`
parameter:

```{r, warning = FALSE, message = FALSE, fig.height=7, fig.width=8}
plot_bw_bins_violin(
  c(h33_chip, h3k9me3_chip),
  bin_size = 50000,
  labels = c("H3.3", "H3K9me3"),
  remove_top = 0.001,
  highlight = genes
)
```

Here you see 5 bins that overlap with the loci in the `genes` file. 

You could also want to pairwise compare the bins. You can do so by plotting
a scatterplot:

```{r, warning = FALSE, message = FALSE, fig.height=7, fig.width=8}
plot_bw_bins_scatter(
  h33_chip,
  h3k9me3_chip,
  bin_size = 50000,
  remove_top = 0.001,
  highlight = genes
)
```

In fact, you may be interested in looking at the signal just at the genes instead:

```{r, warning = FALSE, message = FALSE, fig.height=7, fig.width=8}
plot_bw_loci_scatter(
  h33_chip,
  h3k9me3_chip,
  loci = genes
)

```

You can also check how these two signal behave globally in a more meaningful
way, i.e. in relation with genomic annotations. You can do this with 
`bw_loci_summary_heatmap`:

```{r, warning = FALSE, message = FALSE, fig.width = 11, fig.height = 5}
# ChromHMM is a genome-wide annotation according to epigenetics marks. Each
# locus is tagged by a category. And the amount of categories must be limited.
# In this case, it is fifteen. 
chrom_values <- import(chromhmm, format = "BED")
head(chrom_values)

plot_bw_loci_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  loci = chromhmm,
)
```

In a more detailed way, you can look at the signal profile at the given genes,
using `plot_bw_profile`:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input")
)
```

Still, these profiles are an average of 28 loci. So you may be interested in 
looking at the individual profiles, which you can do with a heatmap view:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes
)
```

In a nutshell, these are the type of things you can do with `wigglescout`. Each
of these functions have many parameters that allow you to fine tune the 
results. You can look at that in more detail in the section below. 

Functionality to calculate the values without plotting them is also provided,
so if you want to plot something different that is not provided as an out-of-the-box
function, you can still use this library for that.

# Calculating functions

bigWig data can be summarized genome-wide using calculation functions: `bw_bins`
`bw_bed`, `bw_heatmap` or `bw_profile`.



## Genome-wide analysis

`bw_bins` function returns a `GRanges` object where each entry is a locus of 
length `bin_size` and its score is the mean coverage.

**Note:** Small bin sizes (< 5000bp) take longer to run.

```{r, warning = FALSE, message = FALSE}
# Several bwfiles can be provided at once
bw_bins(c(h33_chip, input_chip),
        bin_size = 50000,
        genome = "mm9",
        selection = locus)
```

It is also possible to provide a list of bigWig files to be used as background
to normalize the bin values to. For instance, in the previous case, one could
want to use the input values to normalize the H3.3 ChIP data:

```{r, warning = FALSE, message = FALSE}
bw_bins(h33_chip, bg_bwfiles = input_chip,
        bin_size = 50000,
        genome = "mm9",
        selection = locus)
```
If you want to get the log2 ratio sample / input, select `norm_mode = "log2fc"`:

```{r, warning = FALSE, message = FALSE}
bw_bins(h33_chip, bg_bwfiles = input_chip,
        bin_size = 50000,
        genome = "mm9",
        selection = locus,
        norm_mode = "log2fc")
```

`norm_mode` is set to `"fc"` by default, which means it shows fold change 
sample / input, if input (background) is provided.

### Supported genomes

`bw_bins` function is based on `tileGenome` function from `GenomicRanges`, which
requires information about chromosome length. This is obtained relying on
function `Seqinfo` from `GenomeInfoDb` package. So, in theory, any genome ID
that would return a `Seqinfo` object by doing `Seqinfo(genome="your_genome")`
should work.

## Locus-based analysis

These functions are useful when you have some genomic annotations you want to
look at, rather than general genome-wide trends.

### Individual loci

To calculate values on a *locus* specific manner you can use `bw_bed` function.
For instance, we can look at the genes at this region:

```{r, warning = FALSE, message = FALSE}
bw_bed(c(h33_chip, input_chip), bedfile = genes)
```

Many of the options available for `bw_bins` are shared with `bw_bed`, so you can
also use background bigWig files:

```{r, warning = FALSE, message = FALSE}
bw_bed(h33_chip, bg_bwfiles = input_chip, bedfile = genes)
```

A name column is provided if the given `BED` file includes *loci* IDs.

You can also provide `GRanges` objects instead of `BED` files:

```{r, warning = FALSE, message = FALSE}
genes_granges <- rtracklayer::import(genes, format = "BED")

bw_bed(h33_chip, bg_bwfiles = input_chip, bedfile = genes_granges)
```


### Aggregated loci

It is also possible to aggregate values by some kind of category. This is 
useful for `BED` files where ID is not an individual entity but rather a 
category, for instance, ChromHMM annotations:

```{r, warning = FALSE, message = FALSE}
chrom_ranges <- import(chromhmm)

chrom_ranges
```

It is possible to get a single value per category using `aggregate_by` parameter.

```{r, warning = FALSE, message = FALSE}
# If aggregate_by is not provided, you get a non-aggregated GRanges
bw_bed(h33_chip,
       bg_bwfiles = input_chip,
       bedfile = chrom_ranges,
       aggregate_by = "mean") 
```

**Note:** In this case you will not get a `GRanges` object but a `data.frame`.

For these calculations you **always** get some kind of **average** per locus. 
However, different ways of treating the average values are available by changing
the `aggregate_by` parameter:

- `true_mean`: This calculates a mean adding up all the values included per *locus*. 
It is most meaningful when loci represent categories rather than biological
entities.
- `mean`: This calculates a mean-of-means instead, where each *locus* mean value
is calculated and the mean of this distribution of means is returned as a result.
- `median`: This calculates a median-of-means. It is useful to perform an analysis
that is more robust to outliers, but it does not work well with sparse data.

### Profile averages

You may be interested in the **shape** the signal takes across a set of *loci*.
This is calculated by `bw_profile`. This function takes each *locus* in the 
provided bed file and slices it into a fixed number of bins (this is determined
by `bin_size` parameter). Each bin is aggregated together across loci.


```{r, warning = FALSE, message = FALSE}
profile <- bw_profile(h33_chip, bedfile = genes, bin_size = 100)

head(profile)
```

**Note:**. `bw_profile` returns data as long format, since it returns more than
one value per data point. You get: mean, standard error, median value per point.
Each point is represented by index. This index determines its bin.

You can also provide `upstream` and `downstream` base pairs values to include
in this calculation.

```{r, warning = FALSE, message = FALSE}
profile <- bw_profile(h33_chip, bedfile = genes, bin_size = 100,
                      upstream = 500, downstream = 500)

head(profile)
```

Additionally, you can use `mode` parameter to define how these locus are aligned
before binning. Available modes are: `start`, `end`, `center` (align the loci around
these points) or `stretch` which anchors features `start` to `end` and stretches
them when length is different.

### Per-locus profiles

This shape created by `bw_profile` function is still an average. So you may be
interested on a per-locus profile value. This is what you get with `bw_heatmap`.
`bw_heatmap` returns a `matrix` where each row is a locus and each column is
a bin. The rest of parameters are analogous to `bw_profile`:

```{r, warning = FALSE, message = FALSE}
mat <- bw_heatmap(h33_chip, bedfile = genes, bin_size = 1000,
                  upstream = 5000, downstream = 5000)

# Note that bw_heatmap returns a list of matrices, so it supports arrays of 
# bwfiles as input as well.
head(mat[[1]])
```

## General shared options

Each `wigglescout` function has a different set of parameters. You can check
the documentation with `?` operator: `?bw_bins`.

Here are a few that are commonly appearing in many of the functions:

- `labels`: This parameter gives name to the value column of the `data.frame` or
`GRanges` object obtained. By default the names correspond to the file names.
- `remove_top`: Exclude a certain top quantile of the returning object. This 
value must be in [0, 1] and by default is 0 (no values removed). Underneath,
threshold value is calculated by `quantile` function and used as a threshold,
where the exact value is included (values <= threshold are returned). 
If the calculated data structure has more than one column (i.e. multiple files
are provided), quantiles are calculated over `rowMeans`. This prioritizes cases
where outliers are such in more than one sample. In the case where `NA` values
are found, these are also dropped if `remove_top > 0`.
- `per_locus_stat`: Aggregating function per locus. This is an interface with
the underlying `rtracklayer` summary function. It defaults to `mean` and it is
not recommended to change it.


# Plotting functions

Plots can be made with functions:
- `plot_bw_bins_scatter`. A genome-wide bin-based scatterplot. Needs `genome`
and `bin_size` parameters. 
- `plot_bw_loci_scatter`. A locus-based scatterplot. Needs a `loci` file or `GRanges` object.
- `plot_bw_bins_violin`. A genome-wide bin-based violinplot. Needs `genome` and `bin_size` parameters.
- `plot_bw_loci_summary_heatmap`. Summarized heatmap over a categorized BED file.
This works great with annotations like ChromHMM, but other BED files can be used,
as long as their name field is meaningful (provides a category). An example could
be a `genes.bed` file where genes are labelled according to expression, low, med, high.
- `plot_bw_profile`. A profile plot over a set of loci. 
- `plot_bw_heatmap`. A per-locus heatmap.

As you may have noticed, `loci` functions take as parameter a `loci` value, which
can be a BED file or a `GRanges` object. If you provide a `GRanges` object,
chances are that you need to provide a `label`, as there is no way to infer
this from the `GRanges` object. On the other hand, whenever bigWig or BED files
are provided, they are labeled by default with their names.

You can find more detailed information about these below.

## Genome-wide plots

Bin-based genome-wide analysis is currently supported in two formats: 

- `plot_bw_bins_violin`. Makes a violin plot of one or more bigWig files.
- `plot_bw_bins_scatter`. Makes a scatter plot of two different bigWig files
against each other. 

Both plot types allow for background bigWig signal to be provided (i.e. input).

### Violin plots

For a set of bigWig files, it is possible to visualize the genome-wide signal
distribution. For example:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_violin(h33_chip, bin_size = 50000)
```

Note that `bin_size` parameter influences the time any of these functions take
to run. Reasonable resolution for this is 5000 or 10000.

It is possible to plot several bigWig files at a time:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_violin(c(h33_chip, input_chip), bin_size = 50000)
```

Additionally, one can highlight bins that overlap with a certain set of interest
loci. This is done via the `highlight` parameter:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_violin(
  c(h33_chip, input_chip),
  bin_size = 50000,
  highlight = genes,
  highlight_color = c("red", "red") # It is possible to provide different colors
)
```

It is also possible to provide a list of bigWig files to be used as background
to normalize the bin values to. For instance, in the previous case, one could
want to use the input values to normalize the H3.3 ChIP data:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
# bg_bwfiles parameter is set now to input_chip
plot_bw_bins_violin(
  h33_chip, 
  bg_bwfiles = input_chip,
  bin_size = 50000,
  highlight = genes,
  highlight_color = c("red")
)
```

If `norm_mode` is provided, it is possible to log-transform these values as
well:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
# bg_bwfiles parameter is set now to input_chip
plot_bw_bins_violin(
  h33_chip, 
  bg_bwfiles = input_chip,
  bin_size = 50000,
  highlight = genes,
  highlight_color = c("#000088"), # Colors can be also HTML codes
  norm_mode = "log2fc"
)
```

### Genome-wide scatterplots

You can compare two signals against each other on a scatter plot. To
do so, you can use `plot_bw_bins_scatter`. This function works very similarly to
`plot_bw_bins_violin` but instead of accepting an array of bigWig files, it
accepts separate `x` and `y` parameters and the corresponding optional 
`bg_x` and `bg_y`, `norm_mode_x`, `norm_mode_y`. You can see more details in
the function documentation.

One example would be looking at H3.3 versus H3K9me3 in this sample data:

```{r, warning = FALSE, message = FALSE}
plot_bw_bins_scatter(
  x = h33_chip,
  y = h3k9me3_chip,
  bin_size = 50000
)
```

This is useful to see relationships between marks. In this case, bins where
H3.3 signal is high, H3K9me3 signal is low and viceversa.

The same way as with `plot_bw_bins_violin`, it is possible to specify a set of
loci to highlight within the bins, and normalize to given signals:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_scatter(
  x = h33_chip,
  bg_x = input_chip,
  y = h3k9me3_chip,
  bg_y = input_chip,
  bin_size = 50000,
  highlight = genes,
  highlight_colors = "red",
  highlight_label = "genes", # It is also possible to label the groups
  norm_mode_x = "log2fc" # Each norm mode is provided separately
)
```

## Locus-specific plots

It is also possible to look at the ChIP data in a more specific way. For this,
`elsasserlib` provides the following functions:

- `plot_bw_bed_summary_heatmap`: Summarizes the values across a BED file, using
name field as category. 
- `plot_bw_profile`: Creates a profile plot across a set of loci.
- `plot_bw_heatmap`: Creates a heatmap across a set of loci.

### Summary heatmaps

These plots are meant to provide a genome-wide overview of the signal across
different biologically-relevant categories. They get a list of bigWig files
and a BED file as input. Each bigWig file will represent a row in the heatmap,
and each cell will be an aggregate value of the corresponding bigWig file in
a given category in the BED file, for example:

```{r, warning = FALSE, message = FALSE, fig.width=10}
# ChromHMM is a genome-wide annotation according to epigenetics marks. Each
# locus is tagged by a category. And the amount of categories must be limited.
# In this case, it is fifteen. 
chrom_values <- import(chromhmm, format = "BED")
head(chrom_values)

plot_bw_loci_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  loci = chromhmm,
)
```

Since this example is reduced to a set of HOXC genes, you can see how H3.3
signal is high and H3K9me3 is generally low.

It is possible as in the previous types of plots to provide bigWig files as
normalization / input tracks and use them in the same way as before:

```{r, warning = FALSE, message = FALSE, fig.width=10}
plot_bw_loci_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  bg_bwfiles = c(input_chip, input_chip),
  loci = chromhmm
)
```

#### How the values are aggregated

Since this function is aggregating the values, it is possible to specify the 
way the per-locus values are aggregated. Default is **true mean**. What it means
is that all the coverage values are binned together and the mean is performed
as if each category is one big locus. Another possibilities are:
**mean of means** (`aggregate_by = "mean"`), which will calculate mean coverage
per locus and then the aggregate value will be the mean of those mean values,
and **median of means** (`aggregate_by = "median"`). Median per-locus coverage
values tend to be noisy, so a median of means could be useful in some cases.

```{r, warning = FALSE, message = FALSE, fig.width=10}
plot_bw_loci_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  bg_bwfiles = c(input_chip, input_chip),
  loci = chromhmm,
  aggregate_by = "median"   # Note this is a string value, not a function
)
```

### Profile plots

Another way to look at how the ChIP data looks is to summarize a profile across
a set of *loci*. In order to do that you can use the `plot_bw_profile` function:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input")
)
```

The default mode is called `stretch`, where start (3' end) and end (5' end) of
*loci* are aligned and the length of the *locus* is stretched to a given 
pseudolength (`middle` parameter). If this parameter is not given, the pseudolength
will be assigned as the median value of the length distribution of *loci*, 
in an attempt to match a realistic length. 

It is possible to align only 3' ends, 5' ends or center of loci, similarly to
what formatting text alignment would do:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input"),
  mode = "start"
)
```

It is also possible to specify the upstream and downstream lengths:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input"),
  mode = "center",
  upstream = 3000,
  downstream = 3000
)
```

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input"),
  mode = "end",
  upstream = 7000,
  downstream = 3000,
  show_error = TRUE # Turn error estimates on
)

```

### Heatmaps

Finally, it is possible to look at the profile across a set of *loci* in a 
per-*locus* way by looking at heatmaps. 

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes
)
```

This gene annotation includes some variants of the same genes that overlap a lot,
so some of the rows look very similar or even like duplicates.

Most of the parameters are shared between `plot_bw_profile` and `plot_bw_heatmap`
making it possible to arrange them on top of each other later on, so it is
possible to align them in different ways:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  mode = "start",
  upstream = 2500,
  downstream = 2500
)
```

It is possible to specify `bin_size` to increase or decrease resolution (this
can also be done on profile plots):

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  mode = "start",
  upstream = 2500,
  downstream = 2500,
  bin_size = 20
)
```

At this point, heatmap values are shown in order from maximum mean value across
all the beans to lowest.

By default, colorscale is set to percentiles 0.01 and 0.99 of whole matrix 
range. You can set these values to fixed numbers. This is useful mostly to 
plot together several heatmaps with the same color scale. For example:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  mode = "start",
  upstream = 2500,
  downstream = 2500,
  bin_size = 100,
  zmin = 2,
  zmax = 10
)
```

It is also possible to change the color of the heatmap using any `RColorBrewer`
available palette:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  bin_size = 100,
  cmap = "RdYlBu"
)
```

There is also a `max_rows_allowed` parameter that defaults to 10000. This 
allows to plot large matrices by binning vertically the heatmap and plotting the
underlying mean associated, provided that a screen or a reasonable screen will
not be able to display all the data points anyway. If the number of rows is 
larger than `max_rows_allowed`, rows will be binned in a way that the resulting
plotted matrix will have `max_rows_allowed` height and a warning will be shown.
Since the matrix is not subsampled only, but the underlying values averaged,
this may result in a smoothened effect in the image. This is a similar approach
to how image resizing works.

## Customizing plots

All plotting functions return a `ggplot` object. This means that essentially
you can overlay any other data you want or further customize the results. For example:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
profile_plot <- plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes
)

profile_plot + 
  ggtitle("New title for the plot") + # Add a custom title
  geom_hline(yintercept = 1, linetype="dashed") # Add a horizontal line on y=1
```


# On performance and runtime

All these functions work on genome-wide data, and often you will want to run
these on more than one bigWig file at a time. It is possible to run all of this
in a regular laptop, however if resolution is too high, waiting times will 
raise to minutes and even hours, depending on the amount of files and the 
given resolution.

In a Intel i7 laptop, bins analyses for a single bigWig file in resolution
around 10000bp tend to take a few seconds. 5000bp is still reasonable interactive
time. For plotting under 5000 bp resolution you will need to wait quite some
time and I would recommend running these in a script outside R environment.

Locus-based analyses runtime tends to be smaller since the amount of values to
be calculated is smaller than genome-wide bins. An exception to this are 
ChromHMM plots, since these *are* also genome-wide bins, if only of different
lengths and labeled with categories. Keep this in mind when plotting a large
set of bigWig files. It will take some time as well.

## Using multiple processors

As of version 0.2.0, `wigglescout` core functions support `future` specifications
using `furrr` library. This means you can run the code in multiple R sessions:

```{r, warning = FALSE, message = FALSE}
# You just have to set the plan
library(future)
plan(multisession, workers = 2)

# Then run the functions just the same
bw_bins(c(h33_chip, input_chip),
        bin_size = 50000,
        genome = "mm9",
        selection = locus,
        norm_mode = "log2fc")
```

It is advised against implementing future planning within library function 
according to [`future` documentation](https://www.rdocumentation.org/packages/future/versions/1.21.0/topics/plan):

>Please refrain from modifying the future strategy inside your packages / functions, i.e. do not call plan() in your code. Instead, leave the control on what backend to use to the end user. This idea is part of the core philosophy of the future framework - as a developer you can never know what future backends the user have access to. Moreover, by not making any assumptions about what backends are available, your code will also work automatically with any new backends developed after you wrote your code.

## How to find the best configuration

Parallelization in `wigglescout` setting is based on the fact that operations
on multiple files are done separately and independently of each other. However,
doing this requires some passing of data back and forth, which means that if 
your bigWig files are large (or the results you get from them, for instance,
big `GRanges` objects), or you don't have many, you may suffer the overhead
without getting any benefit from it.

An easy rule of thumb is: swap from `sequential` (which is the default) to
`multisession` if you are running functions on multiple files at once. It seems
like there is benefit to do this pretty much for any number of files larger 
than one.

However, adding more sessions does not seem to help a lot. Best guess is going
for 2 or 4 workers. With 2 workers you already see a considerable reduction
in time.

My expectation is that `multicore` may be faster. However I have not benchmarked
it, as it does not run within Rstudio.



