---
title: "Test data documentation"
author: "Carmen Navarro"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Test data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{wigglescout}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Summary 

This is a small document describing the data used to test `wigglescout`
functionality. Since most of the functions look at genomic intervals, 
the files provided are BED files and are included in the package.

# Files content

The provided files are two differently scored BED files and corresponding
mock background files (i.e. input files) to test calculations. These contain
two chromosomes `chr1` and `chr2` of length 200bp each, where signal goes from
1 to 20 and from 20 to 1 on each.

Background files have flat signals of 1 and 2 for the same intervals.

Additionally, there is an annotation file for two types of regions, to validate
summarizing and aggregation across different length intervals.

This dataset will be extended if functionality grows.

```{r setup, message=FALSE, warning=FALSE}
library(wigglescout)
library(rtracklayer)

ranges <- import(system.file("testdata", "bed1.bed", package="wigglescout"))
ranges

```


```{r session-info}
sessionInfo()
```

