---
title: "Wigglescout"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wigglescout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 6
)
```

# Summary and scope

`wigglescout` is an R library that allows you to calculate summary values 
across bigWig files and BED files and visualize them in a genomics-relevant
manner. It is based on broadly used libraries such as `rtracklayer` and
`GenomicRanges`, among others for calculation, and mostly `ggplot2` for
visualization. You can look at the `DESCRIPTION` file to get more information
about all the libraries that make this one possible. 

There are also many other tools whose functionality overlaps a little or
much with `wigglescout`, but there was no single tool that 
included all that I needed. The aim of this library is therefore not to 
replace any of those tools, or to provide a silver-bullet solution to genomics
data analysis, but to provide a comprehensive, yet simple enough set of tools
focused on bigWig files that can be used entirely from the R environment
without switching back and forth across environments.

Other tools and libraries for akin purposes that you may be looking for include:
deepTools, SeqPlots, bwtools, wiggletools, and the list is endless!

`wigglescout` allows you to summarize and visualize the contents of 
bigWig files in two main ways:

- **Genome-wide**. Genome is partitioned on equally-sized bins and
their aggregated value is calculated. Useful to get a general idea of the
signal distribution without looking at specific places.
- **Across sets of *loci***. This can be either summarized categories, or
individual values, as in genome-wide analyses.

`wigglescout` functionality is built in two layers. Names of functions that
calculate values over bigWig files start with `bw_`. These return `GRanges`
objects when possible, `data.frame` objects otherwise (i.e. when values are
summarized over some category, genomic location is lost in this process).

On the other hand, functions that plot such values and that usually make
internal use of `bw_` functions, start with `plot_`.

# About bundled data

This package comes with a set of small files to show functionality. These
have been built from a published data from Simon Elsässer's lab and correspond
to a H3.3 and H3K9me3 ChIP + input (`GSE149080`), subset across a
500kbp genomic region: `chr15-102600000-103100000`, which overlaps with the
*HOXC* gene cluster. A ChromHMM annotation has also been subset to overlap 
with such region.

```{r setup}
library(ggplot2)
library(rtracklayer)
library(GenomicRanges)
library(wigglescout)

h33_chip <- system.file("extdata", "sample_H33_ChIP.bw", package = "wigglescout")
h3k9me3_chip <- system.file("extdata", "sample_H3K9me3_ChIP.bw", package = "wigglescout")
input_chip <- system.file("extdata", "sample_Input.bw", package = "wigglescout")
genes <- system.file("extdata", "sample_genes_mm9.bed", package = "wigglescout")
chromhmm <- system.file("extdata", "sample_chromhmm.bed", package = "wigglescout")

locus <- GRanges(seqnames = "chr15", IRanges(102600000, 103100000))
```

All these values are paths to bigWig and BED files.

# Quick start

Here is a small example of what you can do with `wigglescout`. Imagine you want
to take a look at how values are distributed overall in your bigwig files,
genome-wide:

```{r}
plot_bw_bins_violin(
  c(h33_chip, h3k9me3_chip),
  bin_size = 2000,
  labels = c("H3.3", "H3K9me3"),
  remove_top = 0.001,
  selection = locus # Plot can be subset to a certain GRanges.
)
```

For transparency and reproducibility, plots show relevant underlying values by
default, such as parameters used and calculated values: how many `NA` values were
found, if any, how many points were excluded from the plot due to quantile
cutoff, and so on. Note also that in this case the quantile cutoff seems low,
but it is because top elements are removed according to mean in all samples.

Note also that the bin size chosen, `2000` is somewhat small for genome-wide
analyses (it will take long runtime, see the "Performance and runtime" section
at the end of the document for more details), but it can be done in this case
due to the use of `selection` parameter, which restricts the bin analysis to
a certain locus.

You could also be interested in checking out how are certain *loci*  behaving
in comparison with the global distribution. You can do so providing a `highlight`
parameter:

```{r}
plot_bw_bins_violin(
  c(h33_chip, h3k9me3_chip),
  bin_size = 2000,
  labels = c("H3.3", "H3K9me3"),
  remove_top = 0.001,
  highlight = genes,
  selection = locus # Plot can be subset to a certain GRanges.
)
```

Here you see highlighted the bins that overlap with the loci in the
`genes` file. 

You could also want to pairwise compare the bins. You can do so by plotting
a scatterplot:

```{r}
plot_bw_bins_scatter(
  h33_chip,
  h3k9me3_chip, 
  bin_size = 2000,
  remove_top = 0.001,
  highlight = genes,
  highlight_label = "Genes",
  selection = locus # Plot can be subset to a certain GRanges.
)
```

In real cases where you are plotting the full genome-wide set of bins, which can
be in the tens or hundreds of thousands data points, there can be a lot of 
overplotting and it might be difficult to see where the majority of those are.
For this, you can just plot a 2d histogram instead:

```{r}
plot_bw_bins_density(
  h33_chip,
  h3k9me3_chip,
  bin_size = 20000,
  plot_binwidth = 0.05
)
```


You may be interested in looking at the signal just at the genes instead:

```{r}
plot_bw_loci_scatter(
  h33_chip,
  h3k9me3_chip,
  loci = genes
)

```

You can also check how these two signal behave globally in a more meaningful
way, i.e. in relation with genomic annotations. You can do this with 
`bw_loci_summary_heatmap`:

```{r}
# ChromHMM is a genome-wide annotation according to epigenetics marks. Each
# locus is tagged by a category. And the amount of categories must be limited.
# In this case, it is fifteen. 
chrom_values <- import(chromhmm, format = "BED")
head(chrom_values)

plot_bw_loci_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  loci = chromhmm,
  labels = c("H33", "H3K9me3")
)
```

In a more detailed way, you can look at the signal profile at the given genes,
using `plot_bw_profile`:

```{r}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  loci = genes,
  labels = c("H3.3", "H3K9me3", "Input")
)
```

Still, these profiles are an average of 28 loci. So you may be interested in 
looking at the individual profiles, which you can do with a heatmap view:

```{r}
plot_bw_heatmap(
  h33_chip,
  loci = genes
)
```

In a nutshell, these are the type of things you can do with `wigglescout`. Each
of these functions have many parameters that allow you to fine tune the 
results. You can look at that in more detail in the section below. 

Functionality to calculate the values without plotting them is also provided,
so if you want to plot something different that is not provided as an out-of-the-box
function, you can still use this library for that.

# On performance and runtime

All these functions work on genome-wide data, and often you will want to run
these on more than one bigWig file at a time. It is possible to run all of this
in a regular laptop, however if resolution is too high, waiting times will 
raise to minutes and even hours, depending on the amount of files and the 
given resolution.

In a Intel i7 laptop, bins analyses for a single bigWig file in resolution
around 10000bp tend to take a few seconds. 5000bp is still reasonable interactive
time. For plotting under 5000 bp resolution you will need to wait quite some
time and I would recommend running these in a script outside R environment.

Locus-based analyses runtime tends to be smaller since the amount of values to
be calculated is smaller than genome-wide bins. An exception to this are 
ChromHMM plots, since these *are* also genome-wide bins, if only of different
lengths and labeled with categories. Keep this in mind when plotting a large
set of bigWig files. It will take some time as well.

## Using multiple processors

As of version 0.2.0, `wigglescout` core functions support `future` specifications
using `furrr` library. This means you can run the code in multiple R sessions:

```{r}
# You just have to set the plan
library(future)
plan(multisession, workers = 2)

# Then run the functions just the same
bw_bins(c(h33_chip, input_chip),
        bin_size = 50000,
        genome = "mm9",
        selection = locus,
        norm_mode = "log2fc")
```

It is advised against implementing future planning within library function 
according to [`future` documentation](https://www.rdocumentation.org/packages/future/versions/1.21.0/topics/plan):

>Please refrain from modifying the future strategy inside your packages / functions, i.e. do not call plan() in your code. Instead, leave the control on what backend to use to the end user. This idea is part of the core philosophy of the future framework - as a developer you can never know what future backends the user have access to. Moreover, by not making any assumptions about what backends are available, your code will also work automatically with any new backends developed after you wrote your code.

## How to find the best configuration

Parallelization in `wigglescout` setting is based on the fact that operations
on multiple files are done separately and independently of each other. However,
doing this requires some passing of data back and forth, which means that if 
your bigWig files are large (or the results you get from them, for instance,
big `GRanges` objects), or you don't have many, you may suffer the overhead
without getting any benefit from it.

An easy rule of thumb is: swap from `sequential` (which is the default) to
`multisession` if you are running functions on multiple files at once. It seems
like there is benefit to do this pretty much for any number of files larger 
than one.

However, adding more sessions does not seem to help a lot. Best guess is going
for 2 or 4 workers. With 2 workers you already see a considerable reduction
in time.

My expectation is that `multicore` may be faster. However I have not benchmarked
it, as it does not run within Rstudio.

```{r session-info}
sessionInfo()
```


