---
title: BigWig analysis report template
output: 
  html_document:
    toc: true
    theme: cerulean
    highlight: pygments
params:
  datadir: './bw_test'
  binsize: 10000
  bed: NA
  highlight: NA
  minoverlap: 2000
  genome: 'mm9'
  force: FALSE
  outdir: '.'
---

```{r setup, include=FALSE, messages=FALSE}
library(elsasserlib)
library(ggplot2)
library(reshape2)
library(ggpubr)
library(stringr)
library(pheatmap)
library(RColorBrewer)

# This will be added as functions, but right now is contained in the template, but hidden.
compute.lims <- function(mat) {
  maxmat <- mat
  maxmat[is.na(maxmat)] <- -Inf
  maxval <- max(maxmat)

  minmat <- mat
  minmat[is.na(minmat)] <- Inf
  minval <- min(minmat)

  nsteps <- 21.0

  breaklim <- ceiling(abs(max(abs(c(minval, maxval)))))

  # Compute a reasonable amount of steps?
  stepsize <- 2*breaklim / nsteps

  breakslist <- seq(-breaklim, +breaklim, by=stepsize)
  breakslist
}

summary_heatmap <- function(values, title, size=35) {
  bcolor <- "white"

  breakslist <- compute.lims(values)
  palette <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(length(breakslist))
 
  cellsize <- size
  
  pheatmap(values,
           main=title,
           cellwidth=cellsize,
           cluster_rows=F,
           cluster_cols=F,
           cellheight=cellsize,
           border_color=bcolor,
           breaks=breakslist,
           color=palette,
           display_numbers=TRUE)
}


```

This is a self-explanatory template markdown file that does some default
analysis on a set of `bigWig` files contained in a directory.

```{r parameters, include=FALSE}
# This lists all the files in the path you put there
bwfiles <- list.files(params$datadir)
bwfiles.full <- paste(params$datadir, bwfiles, sep='/')
bed <- params$bed
highlight <- params$highlight
```

## Run summary

This is a summary of the parameters with wich `render_bw_report()` was called.
It is meant as a trace of which data was used to produce this.

- **Data directory**: `r normalizePath(params$datadir)`.
- **BigWig files found**: `r bwfiles`.
- **BED region file**: `r normalizePath(params$bed)`.
- **Highlight file**: `r normalizePath(params$highlight)`.
- **Genome**: `r params$genome`.
- **Bin size**: `r format(params$binsize, scientific=F)`.
- **Force**: `r params$force`.

## Parameters description

These parameters are present on the header of the `Rmd` file. You generate a
report like this calling `render_bw_report()`. See documentation to see how
to do this:

- `datadir`: Where the `bigWig` files are.
- `binsize`: Size of the bin for binned plots.
- `outdir`: Where to put the output files (basically the intermediate tables).
- `bed`: A `BED` file containing relevant regions. This will typically be: 
    i) ChromHMM partition of the genome, 
    ii) Set of repetitive elements, 
    iii) Gene types (highly/low expressed).

    Any `BED` file that contains categories on the name field (4th column) will
    work for such plots.
- `highlight`: If there is a set of loci that you want highlighted in the plots,
    meaning that bins that overlap with any of this will be highlighted in the 
    relevant plots. You can leave this blank for no highlight.
- `minoverlap`: If `highlight` is set, this controls the amount of overlap required
    to consider a bin highlighted (this is still not used but it will be soon).
- `genome`: This is needed for the bin analysis. Currently supported genomes
    are `mm9` and `hg38`.
- `force`: If `TRUE` it will re-generate the tables regardless of whether they
    exist.

## Bin analysis

First bins need to be calculated. This is a costful step, so this markdown will
generate tables as a side-product. If they exist, the tables will just be read.

```{r compute_bins}
precomputed_bins <- paste(outdir, paste('bins_', params$binsize, '.csv', sep=''), sep='/')
precomputed_high <- paste(outdir, paste('high_', params$binsize, '.csv', sep=''), sep='/')

# Costful chunk of code. If ran before, avoid re-running it unless force is
# set to true.
if (file.exists(precomputed_bins) && !params$force) {
  bins.values <- read.csv(precomputed_bins, header=T)
  
} else {
  bins.values <- bw_bins(bwfiles.full,
                         bsize=params$binsize,
                         genome=params$genome)
  
  bins.values <- data.frame(bins.values)
  write.csv(bins.values, precomputed_bins, row.names=FALSE)
}

# Precompute also the bins for highlighting, if necessary.

if (file.exists(highlight)) {
  if (file.exists(precomputed_high) && !params$force) {
    highlight.values <- read.csv(precomputed_high, header=T)
  } else {
    highlight.gr <- import(highlight)
    
    highlight.values <- bw_bins(bwfiles.full,
                                bsize=params$binsize,
                                genome=params$genome,
                                selection=highlight.gr)

    highlight.values <- data.frame(highlight.values)
    write.csv(highlight.values, precomputed_high, row.names=FALSE)
  }
}

```


### Global per-sample bin coverage distribution

This shows a violin plot for global bin coverage distribution per sample.
If `highlight` parameter is set, it will show the corresponding overlapping 
bins' values, along with (to be implemented) a p-value and effect size of the
enrichment compared to global.

```{r plot_violin}
bin.id <- c('seqnames', 'start', 'end')

names <- colnames(bins.values)
names <- names[! names %in% c(bin.id, 'width', 'strand')]

melted.highlight <- melt(highlight.values[, c(bin.id, names)], id.vars=bin.id)
melted.bins <- melt(bins.values[, c(bin.id, names)], id.vars=bin.id)

ggplot(melted.bins, aes(x=variable, y=value)) + 
  geom_violin(fill='#cccccc') +
  theme_elsasserlab_screen() +
  geom_jitter(data=melted.highlight,
              aes(x=variable, y=value, color=variable),
              alpha=0.7) +
  # Zoom in, this removes quite some values
  # However, think that the value output as a warning is the sum of all samples
  # not the bins per sample that are removed, this is way smaller
  ylim(0,10) + 
  xlab('') +
  ylab('Mean coverage') +
  theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1))

```


### Scatterplots for selected samples

You can choose to compare some pair of samples against each other (perhaps
replicates, or some other information that would be relevant to have paired).
This is only set here as a stub, as it is not a parameter of the notebook (yet).
But, once rendered as notebook, you can still modify this manually for a more
complete analysis, or add chunks of code to add more x-y comparisons.


```{r plot_scatter}
used.names <- make.names(bwfiles)
highlight.color <- '#20B2AA'
xcol <- used.names[1]
ycol <- used.names[2]

scatter.values <- bins.values[, c(xcol, ycol)]

ggplot(scatter.values, aes_string(x=xcol, y=ycol)) +
  geom_point(color='#cccccc', alpha=0.8) +
  geom_point(data=highlight.values, aes_string(x=xcol, y=ycol), color=highlight.color, alpha=0.8) +
  theme_elsasserlab_screen() +
  xlim(0,50) +
  ylim(0,50) +
  xlab(xcol) +
  ylab(ycol) +
  ggtitle(paste('Bin coverage scatterplot (bsize =', params$binsize, ')', sep='')) +
  theme(legend.position='none')

```


## Summarized bed plot

The summarized ChromHMM-like plots are made by `bw_bed` function:

```{r precompute_bed}
precomputed_summary <- paste(outdir, paste('truemean_',
                                           basename(params$bed),
                                           '.csv', sep=''),
                             sep='/')

# Costful chunk of code. If ran before, avoid re-running it unless force is
# set to true.
if (file.exists(precomputed_summary) && !params$force) {
  summary.values <- read.csv(precomputed_summary, header=T, row.names=1)

} else {
  summary.values <- bw_bed(bwfiles.full,
                           bed=params$bed,
                           aggregate.by='true_mean')

  write.csv(summary.values, precomputed_summary, row.names=TRUE)
}
```


Once the values are calculated we can plot it:

```{r summary_heatmap, fig.width=12}
summary_heatmap(t(summary.values), "Mean coverage per region")
```

